model-understanding

### 2019-05-05 stats/model-understanding.Rmd

---
title: "Model"
author: "M. Bohn"
output: html_document
---

@@

@@

@@

```{r}

model_pred <- sem_know_model %>%
  separate(Parameter, into = c("type", "item","x"))%>%
  filter(type == "modelPrediction")%>%
  separate(value, into = c("age", "pred"),sep = ", ")%>%
  mutate(age = as.numeric(str_remove_all(age, "[(c]")),
         pred = as.numeric(str_remove(pred, "[)]")))%>%
  select(-x)

item_params<- sem_know_model %>%
  separate(Parameter, into = c("parameter", "type","item"), sep = ",")%>%
  filter(parameter == "parameter", type == "items")%>%
  separate(value, into = c("int", "slope"),sep = ", ")%>%
  mutate(int = as.numeric(str_remove_all(int, "[(c]")),
         slope = as.numeric(str_remove(slope, "[)]")))%>%
  gather(parameter, value, int, slope)

global_params <- sem_know_model %>%
  separate(Parameter, into = c("parameter", "type","token"), sep = ",")%>%
  filter(parameter == "parameter", type == "parameters")%>%
    separate(value, into = c("int", "slope"),sep = ", ")%>%
    mutate(int = as.numeric(str_remove_all(int, "[(c]")),
         slope = as.numeric(str_remove(slope, "[)]")))%>%
    gather(parameter, value, int, slope)

global_sigma<- sem_know_model %>%
  separate(Parameter, into = c("parameter", "type","token"), sep = ",")%>%
  filter(parameter == "parameter", type == "sigma")%>%
    separate(value, into = c("sigma_int", "sigma_slope"),sep = ", ")%>%
    mutate(sigma_int = as.numeric(str_remove_all(sigma_int, "[(c]")),
         sigma_slope = as.numeric(str_remove(sigma_slope, "[)]")))%>%
    gather(parameter, value, sigma_int, sigma_slope)

#### summaries

model_pred_summary <- model_pred %>%
  group_by(item,age)%>%
  summarise(mode = estimate_mode(pred),
            uci = hdi_upper(pred),
            lci = hdi_lower(pred))

item_params_summary <- item_params %>%
  group_by(item, parameter)%>%
  summarise(mode = estimate_mode(value),
            uci = hdi_upper(value),
            lci = hdi_lower(value))

global_params_summary <- global_params %>%
  group_by(token, parameter)%>%
  summarise(mode = estimate_mode(value),
            uci = hdi_upper(value),
            lci = hdi_lower(value))

```{r}
ggplot(item_params, aes(x = value))+#, fill = factor(Chain)))+
  geom_density(alpha = 0.3)+
  xlab("Parameter")+
  facet_grid(parameter~item, scales = 'free')+
  theme_few()
```

```{r}

ggplot(global_params, aes(x = value))#, fill = factor(Chain)))+
  geom_density(alpha = 0.3)+
  xlab("Parameter")+
  facet_grid(parameter~token, scales = 'free')+
  theme_few()
```

```{r}
ggplot(global_sigma, aes(x = value))+#, fill = factor(Chain)))+
  geom_density(alpha = 0.3)+
  xlab("Sigma")+
  facet_grid(~parameter, scales = 'free')+
  theme_few()
```

## Plotting semantic knowledge
```{r}
aoa_ratings_all <- read_xlsx(path = "../data/words_aoa_ratings.xlsx", sheet = 1)

aoa_ratings <- aoa_ratings_all %>%
  filter(Word %in% item_params_summary$item )%>%
  mutate(mean_aoa = as.numeric(Rating.Mean),
         item = Word)%>%
  select(item,mean_aoa)

plot_sem_know <- item_params_summary%>%
  select(-uci,-lci)%>%
  spread(parameter, -item) %>%
  left_join(expand.grid(
      age = unique(me_model_data$age),
      item = me_model_data$familiar), .) %>%
  mutate(sem_know = plogis(int + slope * age)) %>%
  left_join(aoa_ratings) %>%
  ungroup()%>%
  mutate(item = fct_reorder(factor(item), mean_aoa))

ggplot(plot_sem_know, aes(x = age+2, y= sem_know, col = item, group = item))+
  geom_line(size = 1)+
  ylab("Semantic knowledge")+
  xlab("Age")+
  ylim(0,1)+
  theme_few()+
  scale_colour_viridis_d(name = "familiar object")
```

# inverse predictions

```{r}
# speak_opt_params <- speak_opt %>%
#   group_by(parameter)%>%
#   summarise(mode = estimate_mode(value),
#             uci = hdi_upper(value),
#             lci = hdi_lower(value))
# 
# inv_model_data <- sem_know_model %>%
#   filter(Parameter != "speaker_optimality") %>%
#   mutate(Parameter = factor(Parameter)) %>%
#   separate(value, into = c("intercept", "slope"), 
#            sep = ", ") %>%
#   mutate(intercept = as.numeric(str_remove_all(intercept, "[(c]")),
#          slope = as.numeric(str_remove(slope, "[)]"))) %>%
#   rename(item = Parameter)%>%
#   group_by(item)%>%
#   summarise(sem_int = estimate_mode(intercept),
#             sem_slope = estimate_mode(slope))%>%
#   group_by(item)%>%
#   expand(sem_int,sem_slope,age = me_model_data$age)%>%
#   mutate(so_int = speak_opt_params%>%filter(parameter == "intercept")%>%pull(mode),
#          so_slope = speak_opt_params%>%filter(parameter == "slope")%>%pull(mode))%>%
#   mutate(speaker_optimality = so_int + so_slope * age,
#          semantic_knowledge = plogis(sem_int + sem_slope * age))

```{r}
invPred <- '
var allData = dataFromR.data

var output = map(function(row){

    var modelPredictions = pragmaticListener({label: "novel_word"}, priorProbs, row.speaker_optimality,row.semantic_knowledge)

    return extend([row.item + "/" + row.age, Math.exp(modelPredictions.score(1))])

}, allData)

output

```{r predictions pragmatic model ex 3 adults}
# inv_model_pred<- webppl(
#   program_code = paste(rsaUtils, rsaModel, invPred , sep='\n'),
#   data =list(data = inv_model_data),
#   data_var = "dataFromR"
# )
# 
# #saveRDS(inv_model_pred, "../saves/inv_model_pred.rds")
# #inv_model_pred <- readRDS("../saves/inv_model_pred.rds")
# 
# inv_predictions <- inv_model_pred %>%
#   separate(`0`, into = c("item", "age"), sep="/")%>%
#   mutate(age = as.numeric(age))%>%
#   rename( me_effect = `1`)%>%
#   left_join(aoa_ratings)%>%
#   ungroup()%>%
#   mutate(item = fct_reorder(factor(item), mean_aoa), 
#          age = age + min(me_data$age_num))
```

```{r}
plot_model_pred <- model_pred_summary%>%
  left_join(aoa_ratings) %>%
  ungroup()%>%
  mutate(item = fct_reorder(factor(item), mean_aoa))

ggplot(plot_model_pred, aes(x=age+2,y = mode, col = item))+
  geom_hline(yintercept = 0.5, lty=2)+
  geom_line(size = 1)+
  ggtitle("Inverse Model Predictions")+
  labs(x="Age",y="Proportion correct")+
  theme_few() +
  ylim(-0.05,1.05)+
  xlim(2,5)+
  guides(alpha = F)+ 
  scale_colour_viridis_d()

## For comparison: GLMM for ME effect and data
```{r}
me_data <- read_csv("../data/child_basic_me.csv")%>%
    filter(trial != "train1",
           trial != "train2")%>%
  mutate(z_age = age_num - mean(age_num))

# me_model <- brm(correct ~ z_age + (1|subid) + (z_age | familiar),
#                 data = me_data,
#                 family = bernoulli(),
#                 control = list(adapt_delta = 0.95),
#                 sample_prior = F,
#                 save_all_pars = TRUE,
#                 iter = 2000)

#saveRDS(me_model, "../saves/me_effect_model.rds")

me_model <- readRDS("../saves/me_effect_model.rds")

summary(me_model)

fixef <- as_tibble(fixef(me_model), rownames = "term")

ranef <-  ranef(me_model)

age_me_effect <- as_tibble(ranef$familiar, rownames = "familiar")%>%
  mutate(grand_intercept = fixef%>%filter(term=="Intercept")%>%pull(Estimate),
         grand_slope = fixef%>%filter(term=="z_age")%>%pull(Estimate))%>%
  group_by(familiar) %>%
  expand(Estimate.Intercept,Estimate.z_age,grand_intercept,grand_slope,z_age = me_data$age_num - mean(me_data$age_num), mean_age = mean(me_data$age_num))%>%
  mutate(me_effect = plogis(grand_intercept + Estimate.Intercept+(Estimate.z_age+grand_slope)*z_age), 
         age = z_age+mean_age,
         item = familiar)%>%
  select(item, me_effect, age)%>%
  left_join(aoa_ratings)%>%
  ungroup()%>%
  mutate(item = fct_reorder(factor(item), mean_aoa))

model_plot <- ggplot(age_me_effect, aes(x=age,y = me_effect, col = item))+
  geom_hline(yintercept = 0.5, lty=2)+
  geom_line(size = 1)+
  ggtitle("GLMM")+
  labs(x="Age",y="Proportion correct")+
  theme_few() +
  ylim(-0.05,1.05)+
  xlim(2,5)+
  guides(alpha = F)+ 
  scale_colour_viridis_d(name = "familiar object")

p2 <- read_csv("../data/child_basic_me.csv")%>%
  filter(trial != "train1",
         trial != "train2")%>%
  mutate(item = familiar)%>%
    left_join(aoa_ratings)%>%
  ungroup()%>%
  mutate(familiar = fct_reorder(factor(item), mean_aoa))

data_plot <- ggplot(data = p2, aes(x = age_num, y = correct, col = familiar, fill = familiar)) +
  geom_hline(yintercept = 0.5, lty=2)+
  geom_jitter(width = 0, height = 0.05, alpha = .6)+
  geom_smooth(method = "loess", se = F, alpha = .5, span = 2)+
  labs(x="Age",y="Mutual Exclusivity effect")+
  #facet_grid(~familiar)+
  ggtitle("Data")+
  theme_few() +
  ylim(-0.05,1.05)+
  xlim(2,5)+
  guides(alpha = F)+ 
  scale_colour_viridis_d(name = "familiar object")+
  scale_fill_viridis_d(name = "familiar object")
data_plot
```

```{r}
ggarrange(data_plot,inv_model_plot,model_plot, common.legend = T, legend = "right", nrow = 1, ncol = 3)
```

#### WORK IN PROGRESS: nesting within participant

to do:
- bookkeeping: index correct participant/item variables at relevant part of code
- inference: try HMC, variational

```{r work in progress}
semStr <- '
var data = dataFromR.data

  // global parameters
  var so_slope = uniformDrift({a: -2, b: 2, width: 0.2})
  var so_int = uniformDrift({a: -2, b: 2, width: 0.2})

  var item_slope_sigma = uniformDrift({a: 0, b: 5, width: 0.2})
  var item_int_sigma = uniformDrift({a: 0, b: 5, width: 0.2})

  var participant_age_adjust = gaussianDrift({mu: 0, sigma: 1, width: 0.2})

  var sem_slope_item = repeat(familiars.length, function(){
    gaussianDrift({mu: sem_slope_global, sigma: item_slope_sigma, width: 0.5} )
  })

  var sem_int_item = gaussianDrift({mu: sem_int_global, 
        sigma: item_int_sigma, width: 0.5})

      query.add(["modelPrediction", cndtn], [age, Math.exp(modelPredictions.score(1))])

    query.add(["parameter", cndtn], [sem_int_item, sem_slope_item])

  query.add(["parameter", "speaker_optimality"], [so_int, so_slope])
  query.add(["parameter", "global_sem"], [sem_int_global, sem_slope_global])
  query.add(["parameter", "global_sem_sigmas"], [item_int_sigma, item_slope_sigma])

>>>>

---
title: "model-understanding"
author: "MH Tessler"
date: "5/4/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Breaking down the model, for understanding

@@

## RSA model

@@

## Priors for speaker optimality

```{r FIXME}
semKnowPriors <- '
var data = dataFromR.data

## Priors for inferring semantic knowledge

```{r old}
semKnowPriors <- '

var levels = function(df, label){
  return _.uniq(_.map(df, label));
}

var data = dataFromR.data

var discreteUniformBins = function(lower, upper, bin_width){
  _.range(lower, upper, bin_width)
}

var all_ages = sort(levels(data, "age"))

  var globalLineParameters = {
    intercept:  discreteUniformBins(0, 5, 2),
    slope: discreteUniformBins(-2, 2, 2)
  }

  var itemVariability = {
    intercept: discreteUniformBins(0.01, 5, 2),
    slope: discreteUniformBins(0.01, 5, 2)
  }
  // conceptually it might not make sense to have two sigmas, but maybe it does
  // intercept sigma probably will be smaller than slope sigma

  // display([
  //  globalLineParameters.intercept, itemVariability.intercept,
  //  globalLineParameters.slope, itemVariability.slope,
  //  itemLineParameters.intercept, itemLineParameters.slope
  // ])
  var output = map(function(globalLineIntercept){
    map(function(globalLineSlope){
      map(function(itemVariabilityIntercept){
          map(function(itemVariabilitySlope){

            var itemLineParameters = {
              intercept: discreteUniformBins(
                globalLineIntercept - 1*itemVariabilityIntercept,
                globalLineIntercept + 1*itemVariabilityIntercept,
                itemVariabilityIntercept
              ),
              slope: discreteUniformBins(
                globalLineSlope - 1*itemVariabilitySlope,
                globalLineSlope + 1*itemVariabilitySlope,
                itemVariabilitySlope
              )
            }

            map(function(itemLineIntercept){
                map(function(itemLineSlope){
                  map(function(age){
                    var sem_knowledge = logistic(itemLineIntercept + 
                      itemLineSlope * age)
                    // display(itemLineParameters.intercept + "+ " + itemLineParameters.slope + " X " + age + " =logistic= " + sem_knowledge)
                    var priorProbs =  [.5,.5]
                    var speakerOptimality = 1

                    var modelPredictions = pragmaticListener({label: "novel_word"}, 
                            priorProbs, speakerOptimality, sem_knowledge)
                    var smk =  "semKnowledge_" + age
                    var mpa =  "modelPrediction_" + age
                    return {
                      global_intercept: globalLineIntercept,
                      global_slope: globalLineSlope,
                      sigma_intercept:itemVariabilityIntercept, 
                      sigma_slope:itemVariabilitySlope, 
                      item_intercept: itemLineIntercept,
                      item_slope: itemLineSlope,
                      age: age,
                      sem_knowledge : sem_knowledge,
                      L1 : Math.exp(modelPredictions.score(1))
                    }

                  }, all_ages)

              }, itemLineParameters.slope)
            }, itemLineParameters.intercept)

        }, itemVariability.slope)
      }, itemVariability.intercept)
    }, globalLineParameters.slope)
  }, globalLineParameters.intercept)

  _.flattenDeep(output)
  // return query
}
'
```

sem_know_model<- webppl(
  program_code = paste(rsaUtils, rsaModel, semKnowPriors , "model()", sep='\n'),
  data = list(data = me_model_data),
  data_var = "dataFromR",
  model_var = "model",
  #inference_opts = list(method = "enumerate")
)

# copied code

=>
```{r}
save(sem_know_model, "parameter_enumeration_so1_flatPriors.RData")
load(file = "parameter_enumeration_so1_flatPriors.RData")# sem_know_model
```

@@
